Domain Specific Languages and Code Synthesis using Haskell

### 2014-03-14 paper.tex
Rest of 4

Expressing function calls in terms of constructors,  and
building expression trees is a useful, but by itself is a gimmick. 
However, with careful construction, we can also capture function definitions,
and other syntactical structures,
directly from a deep embedding. It is at this
point, the idea of capturing code,
the using the captured code to execute code
on a different target becomes possible.
Consider a simple function to add one to its argument.

@@

First we need a structure for our Lava Language. We define the
functions use above, but give then a deep embedding, called \verb|Signal|.
\begin{Code}

@@

\end{Code}
Now, if we attempt to extract \verb|counter|, things go horribly wrong.
\begin{Code}
GHCi> counter (Var "restart") (Var "inc")
Mux2 (Var "inc") (Add (Mux2 (Var "restart") (Lit 0,Register 0 (Mux2 (Var "inc") ...
\end{Code}        
The output tree is infinite.
What has happened is the recursive definitions are unrolling when we try reify the function,
or more specifically, the body of counter is looping. At this point, the EDSL community was
stymied. There was efforts to use monadic structure, where the loop was expressing using
\verb|do|-notation~\cite{Erkok:2000:Recursive-Monad},
making the loop an observable effect. There was an unsafe
extension to observe a limited form of sharing, by circumventing part of the purity of Haskell,
called observable sharing~\cite{Claessen:99:ObserveSharing}.
There was also an extension of Haskell input/output mechanism that allowed loops to
be observed indirect, called IO-based observable sharing~\cite{Gill:09:TypeSafeReification}. 
The net effect of all three
mechanisms is that the observed tree is rendered as a graph with named edges.

@@

Haskell at this point rescues us from some complexity. Advanced type-system mechanisms, such has higher-kind type
arguments, allow a structure to be either a tree or graph, depending on type-level instantiation. 
Omitting here the details, the reified function is a tree with sharing,
then translated into a graph with explicit sharing. The final result for our example is
\begin{Code}
GHCi> reify (counter (Var "restart") (Var "inc"))
[(0,MUX2 1 (2,3)),
 (1,VAR "inc"),
 (2,ADD 3 4),
 (3,MUX2 5 (6,7)),
 (4,LIT 1),
 (5,VAR "restart"),
 (6,LIT 0),
 (7,REGISTER 0 0)]
\end{Code}
In this output, each upper-case constructors correspond to its deep-embedding constructor.
A quick inspection shows that we have captured the circuit as given in Figure~\ref{fig:counter-picture}.
From this netlist-style structure, it is straightforward to generate VHDL. For the example
of 4-bit number, the VHDL is given in Figure~\ref{fig:counter-VHDL}.

These two tricks (prototypical argument, IO-based observable sharing) are the technical fundamentals
of Kansas Lava. On top of this base, and with help from Haskell type system, an entire ecosystem
for circuit generation has been developed. The DSL idiom allows programmers to use high-level 
abstraction, in Haskell, and generate efficient circuits. But not all is rosy; writing a Lava
program is not the same as writing a Haskell program because of the limitations of Deep
embeddings.

>>>>

Expressing function calls in terms of constructors and
building expression trees is a useful, but by itself is a gimmick. 
However, with careful construction, we can also capture function definitions,
and other syntactical structures,
directly from a deep embedding. It is at this
point, the idea of capturing code,
the using the captured code to execute code
on a different target becomes possible.
Consider a simple function to add one to its argument.

@@

First we need a structure for our Lava Language. We define the
functions used above, but give them a deep embedding, called \verb|Signal|.
\begin{Code}

@@

\end{Code}
Now, if we attempt to extract \verb|counter|, things go horribly wrong.
\begin{Code}
GHCi> counter (Var "restart") (Var "inc")
Mux2 (Var "inc") (Add (Mux2 (Var "restart") (Lit 0,Register 0 (Mux2 (Var "inc") ...
\end{Code}        
The output tree is infinite.
What has happened is the recursive definitions are unrolling when we try reify the function,
or more specifically, the body of counter is looping. At this point, the EDSL community was
stymied. There were efforts to use monadic structure, where the loop was expressing using
\verb|do|-notation~\cite{Erkok:2000:Recursive-Monad},
making the loop an observable effect. There was an unsafe
extension to observe a limited form of sharing, by circumventing part of the purity of Haskell,
called observable sharing~\cite{Claessen:99:ObserveSharing}.
There was also an extension of Haskell input/output mechanism that allowed loops to
be observed indirectly, called IO-based observable sharing~\cite{Gill:09:TypeSafeReification}. 
The net effect of all three
mechanisms is that the observed tree is rendered as a graph with named edges.

@@

Haskell at this point rescues us from some complexity. Advanced type-system mechanisms, such has higher-kind type
arguments, allow a structure to be either a tree or graph, depending on type-level instantiation. 
Omitting here the details, the reified function is a tree with sharing,
then translated into a graph with explicit sharing. The final result for our example is
\begin{Code}
GHCi> reify (counter (Var "restart") (Var "inc"))
[(0,MUX2 1 (2,3)),
 (1,VAR "inc"),
 (2,ADD 3 4),
 (3,MUX2 5 (6,7)),
 (4,LIT 1),
 (5,VAR "restart"),
 (6,LIT 0),
 (7,REGISTER 0 0)]
\end{Code}
In this output, each upper-case constructors correspond to its deep-embedding constructor.
A quick inspection shows that we have captured the circuit as given in Figure~\ref{fig:counter-picture}.
From this netlist-style structure, it is straightforward to generate VHDL. For the example
of 4-bit numbers, the VHDL is given in Figure~\ref{fig:counter-VHDL}.

These two tricks (prototypical argument, IO-based observable sharing) are the technical fundamentals
of Kansas Lava. On top of this base, and with help from the Haskell type system, an entire ecosystem
for circuit generation has been developed. The DSL idiom allows programmers to use high-level 
abstraction, in Haskell, and generate efficient circuits. But not all is rosy; writing a Lava
program is not the same as writing a Haskell program because of the limitations of Deep
embeddings.
